Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-07-10 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to identify classes of events in the early stages of the data analysis, thereby improving its efficiency by limiting the computationally expensive processing to events of interest only, which are often not the most abundant.
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We explore also the application of autoencoder neural networks to clustering by implementing. With autoencoder neural networks we find improved descriptions of data from experiments.

@@

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled dataset is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated dataset. In the latter case, there must also exist a validation dataset of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The method used in \cite{Bradt2018} to analyze the data followed a three-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that become prohibitive in data sets larger by an order of magnitude. 
% Begin added by DB (06/10/2020)
The cleaning was performed using a combination of linear and circular Hough transforms on a 2-dimensional projection of the tracks \cite{Bradt2017}. The following filtering and fitting phases were performed simultaneously, by defining the cost function to the fitting algorithm as a sum of three $\chi^2$ components based on i) the position of the track in space, ii) the energy deposited on each pixel of the sensor plane, and iii) the location of the vertex of the reaction. While various fitting algorithms were tested, the most accurate was a Monte-Carlo algorithm that explored the six-dimensional phase space of the particle's kinematics parameters, reducing it progressively at each iteration step until the desired accuracy was reached \cite{Bradt2017}. Although this algorithm ended up being the most accurate, it is extremely costly computationally because of the very large number of simulated tracks needed for each event. The filtering was performed by setting limits to the $\chi^2$ distributions, below which the events were assigned as proton scattering. This is a very inefficient method because it requires performing the fitting for all events, including those that are not of interest. Pioneering work on event identification using machine learning methods, namely the use of a pre-trained convolutional neural networks (CNN), later showed that the filtering phase would better be performed using this type of technique \cite{Kuchera2019}. 
% End added by DB (06/10/2020)
In addition, \cite{Kuchera2019} demonstrated that the purity and statistics of the data are improved with the use of ML.

@@

The AT-TPC data we studied for this work was recorded as charge time-series for each of the  the $\sim10^4$ detector pads.
In this representation, an event is a record of $512$ time-buckets for each of the $10^4$ detector pads. In our analysis, we represent each event as a downsampled 2D projection.
We chose to represent the data in 2D to facilitate the use of sophisticated image-recognition machine learning models, and this data representation was shown to successfully classify the events of this experiment in a supervised setting \cite{Kuchera2019}.
First, the time-series data was represented as a 3D point cloud, where each point contains the maximum charge in the time-series trace. We log-transform then min-max scale the charge, mapping the charge to the interval $[0, 1]$. The data is projected into 2D space by summing over the time-axis, then downsampled into an $128\times128$ pixel image.
%Additionally, we log-transform and normalize the data to the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the best point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for analysis.
%On the opposite end of the pipeline, 
Therefore, incrementally add bias to the analysis by applying the algorithm further down the analysis pipeline.. With the benefit being that more pre-processing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and pre-processing bias, we performed analysis on simulated, cleaned and raw events.

\begin{table}
\centering
\caption{Event class percentages for each of the datasets used in this analysis. The decrease in the "other" class of events from full to filtered data owes to the thresholding of events. If an event contains fewer than $20$ data-points it is discarded }\label{tab:class_distr}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
$\%$ Proton & $50$ & $25.3$ & $27.6$ \\ 
$\%$ Carbon & $50$ & $12.0$ & $11.9$ \\
$\%$ Other & $0$ & $62.8$ & $60.1$ \\
\end{tabular}
\end{table}

@@

A set of $N=40000$ simulated AT-TPC events per class were simulated from a ${}^{46}$Ar$(p, p)$ experiment with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

For validation, we set a subset of the simulated data to be labelled and treat the rest as unlabelled data. We choose this partition to be $15\%$ of each class. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, while the entire dataset which is identified as $\boldsymbol{X}_F$. Note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to note that what we mean is a raw 2D representation of peak-only events.

@@

We display two different events from the ${}^{46}$Ar experiment in figure \ref{fig:samples}. The top row illustrates a carbon event with a large fraction of noise, while the bottom row shows a proton event nearly devoid of noise.

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{custom_work/examples_raw.pdf}
\caption[Displaying unfiltered events in 2D and 3D]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the color intensity of each point indicates higher charge values recorded by the detector.}\label{fig:samples}
\end{figure}

@@

As we saw in the previous section, the detector picks up a significant amount of noise. We split the noise broadly in two categories,  one being random-uncorrelated noise and the second is structured noise. The former can be quite trivially removed with a nearest-neighbour algorithm that checks if a point in the event is close to any other. To remove the correlated noise, researchers at the NSCL developed an algorithm based on the Hough transform \cite{Newman1972}. This transformation is a common technique in computer vision, used to identify common geometric shapes like lines and circles, and has been used extensively in high-energy particle physics since the bubble-chamber era \cite{Hough:1959}.  %Essentially, the algorithm draws many lines (of whatever desired geometry) through each data-point and checks whether these lines intersect with points in the dataset. Locations in parameter space that generate many intersections then become bright spots, allowing us to filter away points that are not close to these points. These algorithms remove a large amount of the unstructured noise and are computationally rather cheap. \todo{Also pretty thesis-y language. Do not like}

@@

\subsection{Classifying  events} 
%The filtering task relied on a set of $\chi^2$ distributions based on different criteria, 
The traditional Monte-Carlo event selection process, described in Section~\ref{Sec:Exp}, does not have a well-defined method to quantify the effectiveness of the event selection.
In addition, the selection task produced a binary result only, ``good" or``bad", relative to the event of interest, while the data clearly contained more classes of events.%, if only because of the presence of carbon in the target (from the isobutane gas).
In a broader perspective, an unsupervised classification algorithm would offer the possibility to ``discover" rare events not anticipated (or overlooked) by the physicist. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely computationally expensive because it involves the simulation of thousands of tracks for each recorded event, which are performed each iteration of the Monte-Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the sheer number of events to analyse (in the hundreds of thousands) necessitated the use of parallel processing on a high performance computing cluster {\color{blue} is there a metric we can give here? -MPK}. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. 
%Most of the CPU time would be spent on simulating events that are not of interest.
The computationally expensive fitting procedure would be applied to every event, instead of the few percent of the events that are of interest for the analysis.
An unsupervised ML algorithm able to separate the data without a priori knowledge of the different types of events increases the efficiency of the analysis tremendously, and allows the downstream analysis to concentrate on the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data.}\label{tab:datasets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

@@

% In the Machine Learning community, it is not uncommon to publish packaged models with fitted parameters from image recognition contests. These models are trained on datasets with millions of images and classify between hundreds of distinct classes; one such is the imagenet dataset. In their work, \citet{Kuchera2019} used the VGG16 architecture trained on imagenet to classify ${}^{46}$Ar data. In our work we also found the VGG16 architecture to capture the strongest event-separating representation for clustering purposes.

@@

Training high-performing neural networks from scratch often requires enormous datasets and computation time. However, it has been found that models which are trained at large scale will learn general features that are applicable to a variety of tasks. For example, large neural networks which are trained on the ImageNet dataset \cite{Russakovsky2015} --- a diverse image classification task --- learn how to identify lines, edges, and other common shapes that are useful for numerous problems. Thus, it is common practice to initialize the convolutional layers of a network with the pre-trained weights learned from ImageNet (or some other large dataset). The training process then only has to fine-tune the network for the specific task. Since we are building on prior knowledge in this case, learning becomes far more efficient, and better performance can often be achieved. \citet{Kuchera2019} used machine learning methods to classify the products of $^{46}$Ar reactions in the AT-TPC, and they found that a CNN initialized with ImageNet weights resulted in the most successful classification.

@@

%We begin by investigating the K-means clustering algorithm on the VGG16 latent space. 
% The user a VGG16 model that is pre-trained on the ImageNet dataset and subsequently applied to the AT-TPC events without the classification layers. This creates a set of vectors $\boldsymbol{x} \in \R^{8192}$ as our new representation of the events. To cluster, we use the \lstinline{scikit-learn} implementation of the K-means++ algorithm with default parameters \cite{Pedregosa2011}.

@@

The performance is measured by comparing model predictions on the labelled subset of data (see table \ref{tab:datasets}).

@@

The results of the clustering runs are included in table \ref{tab:clstr_vgg}. We observe that we are able to attain near-perfect clustering on simulated data and that there is a sharp decline in performance as we add noise by moving to the filtered and raw datasets.

@@

{\color{blue} This section may contain info that you can move to the metrics section??}
In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we tabulate the assignments of samples relative to their ground truth labels. From these tables, we can infer which classes are more or less entangled with others. The results for each dataset is shown in figures \ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. Purity is inferred by how much spread there is in the column between the ground truth labels. A high-quality cluster will, in addition to being pure, also capture most entries the class represented by the cluster. For example, consider the row corresponding to the proton class in figure \ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it. From this, we conclude that the proton cluster is a high quality, high purity cluster.

@@

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{Illustrating a sample of proton events from different K-means clusters from the filtered dataset. Each row belongs to a single cluster corresponding to the filtered confusion matrix in figure \ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{Illustrating a sample of proton events from different K-means clusters from the raw dataset. Each row belongs to a single cluster corresponding to the full confusion matrix in figure \ref{fig:clster_confmati_raw}}\label{fig:full_vgg_clster_repr}
\end{figure}

@@

\section{Conclusions and Perspectives}\label{sec{conclusion}}
% Begin edit by DB (6/17/2020)
The purpose of this study is to explore the application of unsupervised learning algorithms to event identification from an active target detector. The necessity to identify events from raw data prior to full processing is becoming a major issue in the data analysis of detectors with complex responses such as the AT-TPC.
As shown by both avenues explored in this work, it is clear that there is significant potential to eventually achieve event classification using fully automated unsupervised methods.

@@

In summary, our study shows that unsupervised track classification with an implementation of the VGG16 + K-means approach is a viable solution. For future work, it is worth investigating whether the adaptation of bespoke models like the MIXAE algorithm will allow better performance at no significant cost to consistency.  
%Especially considering the flexibility to use 3D input representations or auxiliary physical goals in the optimization procedure. 
The two examples of unsupervised machine learning methods studied in this work are a first encouraging step towards automated classification of events that could not only greatly reduce the resource cost of analysis, but also eventually boost the efficiency of the experiment by allowing post-trigger decisions based on such algorithm implemented in hardware. Much progress remains to be done before achieving these goals.
% End edit by DB (6/17/2020)
%From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. {\color{orange}[flip this to positive ]} It is also clear that there is significant potential in both avenues explored in this work. 
%n particular, the surprising ability of the K-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
%Another surprising facet of the K-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.

@@

\appendix  
\section{VGG16}\label{app:vgg}
The choice of the kernel size is based on the fact that a stacked $3 \times 3$ kernel is equivalent to larger kernels in terms of the receptive field of the output. Three $3 \times 3$ kernels with stride $1$ have a $7 \times 7$ receptive field, but the larger kernel has $81\%$ more parameters and only one non-linearity \cite{Simonyan2014}. Stacking the smaller kernels then contributes to a lower computational cost. Additionally, there is a regularizing effect from the lowered number of parameters and increased explanatory power from the additional non-linearities.
\begin{table}
\caption[vgg architectures]{Showing the details of the VGG network architectures. Network D trained on the ImageNet \cite{Russakovsky2015} dataset the network known as  VGG16 and is what we use in this thesis.}\label{tab:vgg}
\includegraphics[width=\textwidth]{plots/vgg_architectures.png}
\end{table}

@@

To train the MIXAE clustering algorithm, we use the large simulated dataset with $M=80000$ points, evenly distributed between proton- and carbon-events. The algorithm is trained on a subset of $60000$ of these samples, and we track performance on the remaining $20000$ events. Since there are then only three remaining hyperparameters we choose to perform a coarse grid-search as described in section \ref{sec:hyperparam_search_arch}.Finally, for these parameters we re-ran the algorithm $N=10$ times to investigate the stability of the algorithm.

@@

For the full dataset the MIXAE hyperparameters converge to the same values as for the clean data:

>>>>

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to identify classes of events in the early stages of the data analysis, thereby improving its efficiency by limiting the computationally expensive processing to events of interest only, which are often not the most abundant.
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We explore also the application of autoencoder neural networks to clustering by implementing. With autoencoder neural networks we find improved descriptions of data from experiments.

@@

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled data set is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated data set. In the latter case, there must also exist a validation data set of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The method used in \cite{Bradt2018} to analyze the data followed a three-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that become prohibitive in data sets larger by an order of magnitude.  In particular, as for the data presented and analyzed here, with data sets in the terabyte region or larger, these more traditional methods become prohibitively expensive from a computational stand.
% Begin added by DB (06/10/2020)
The cleaning was performed using a combination of linear and circular Hough transforms on a 2-dimensional projection of the tracks \cite{Bradt2017}. The following filtering and fitting phases were performed simultaneously, by defining the cost function to the fitting algorithm as a sum of three $\chi^2$ components based on i) the position of the track in space, ii) the energy deposited on each pixel of the sensor plane, and iii) the location of the vertex of the reaction. While various fitting algorithms were tested, the most accurate was a Monte-Carlo algorithm that explored the six-dimensional phase space of the particle's kinematics parameters, reducing it progressively at each iteration step until the desired accuracy was reached \cite{Bradt2017}. Although this algorithm ended up being the most accurate, it is extremely costly computationally because of the very large number of simulated tracks needed for each event. The filtering was performed by setting limits to the $\chi^2$ distributions, below which the events were assigned as proton scattering. This is a very inefficient method because it requires performing the fitting for all events, including those that are not of interest. Pioneering work on event identification using machine learning methods, namely the use of a pre-trained convolutional neural networks (CNN), later showed that the filtering phase would better be performed using this type of technique \cite{Kuchera2019}. 
% End added by DB (06/10/2020)
In addition, the authors of Ref.~\cite{Kuchera2019} demonstrated that the purity and statistics of the data are improved with the use of ML.

@@

The AT-TPC data we studied for this work was recorded as charge time-series for each of the  the $\sim10^4$ detector pads.
In this representation, an event is a record of $512$ time-buckets for each of the $10^4$ detector pads. In our analysis, we represent each event as a down-sampled two-dimensional  projection.
We chose to represent the data in two dimensions to facilitate the use of sophisticated image-recognition machine learning models, and this data representation was shown to successfully classify the events of this experiment in a supervised setting \cite{Kuchera2019}.
First, the time-series data was represented as a three-dimensional  point cloud, where each point contains the maximum charge in the time-series trace. We log-transform then min-max scale the charge, mapping the charge to the interval $[0, 1]$. The data is projected into two-dimensional  space by summing over the time-axis, then down-sampled into a $128\times128$ pixel image.
%Additionally, we log-transform and normalize the data to the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the best point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for analysis.
%On the opposite end of the pipeline, 
Therefore, we incrementally add bias to the analysis by applying the algorithm further down the analysis pipeline, with the benefit being that more preprocessing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and preprocessing bias, we performed our analysis on simulated, cleaned and raw events.

\begin{table}
\centering
\caption{Event class percentages for each of the data sets used in this analysis. The decrease in the "other" class of events from full to filtered data owes to the thresholding of events. If an event contains fewer than $20$ data points it is discarded }\label{tab:class_distr}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
$\%$ Proton & $50$ & $25.3$ & $27.6$ \\ 
$\%$ Carbon & $50$ & $12.0$ & $11.9$ \\
$\%$ Other & $0$ & $62.8$ & $60.1$ \\
\end{tabular}
\end{table}

@@

A set of $N=40000$ simulated AT-TPC events per class was obtained  from a ${}^{46}$Ar$(p, p)$ experiment with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

For validation, we set a subset of the simulated data to be labeled and treat the rest as unlabeled data. We chose this partition to consist of  $15\%$ of the data. We denote this subset and its associated labels as $\gamma_L=(\boldsymbol{X}_L, \boldsymbol{y}_L)$, while the entire data set is identified as $\boldsymbol{X}_F$. Note that $\boldsymbol{X}_L \subset \boldsymbol{X}_F$.

@@

The events analyzed in this section were retrieved from the ${}^{46}$Ar resonant proton scattering experiment recorded with the AT-TPC. 
While we denote these events as raw, it is important to note that what we mean is a raw two-dimensional representation of peak-only events.

@@

We display two different events from the ${}^{46}$Ar experiment in figure \ref{fig:samples}. The top row illustrates a carbon event with a large fraction of noise, while the bottom row shows a proton event almost free of noise.

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{custom_work/examples_raw.pdf}
\caption[Displaying unfiltered events in two and three dimensions]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the color intensity of each point indicates higher charge values recorded by the detector.}\label{fig:samples}
\end{figure}

@@

As we saw in the previous section, the detector picks up a significant amount of noise. We split the noise broadly in two categories,  one being random-uncorrelated noise and the second is structured noise. The former can be quite trivially removed with a nearest-neighbour algorithm that checks if a point in the event is close to any other. To remove the correlated noise, researchers at the NSCL developed an algorithm based on the Hough transform \cite{Newman1972}. This transformation is a common technique in computer vision, used to identify common geometric shapes like lines and circles, and has been used extensively in high-energy particle physics since the bubble-chamber era \cite{Hough:1959}.  %Essentially, the algorithm draws many lines (of whatever desired geometry) through each data-point and checks whether these lines intersect with points in the data set. Locations in parameter space that generate many intersections then become bright spots, allowing us to filter away points that are not close to these points. These algorithms remove a large amount of the unstructured noise and are computationally rather cheap. \todo{Also pretty thesis-y language. Do not like}

@@

\subsection{Classifying  events} 
%The filtering task relied on a set of $\chi^2$ distributions based on different criteria, 
The traditional Monte-Carlo event selection process, described in Section~\ref{Sec:Exp}, does not have a well-defined method to quantify the effectiveness of the event selection.
In addition, the selection task produced a binary result only, either ``good" or``bad" relative to the event of interest. The data contain however many more classes of events.%, if only because of the presence of carbon in the target (from the isobutane gas).
In a broader perspective, an unsupervised classification algorithm would offer the possibility to ``discover" rare events which may not be expected or are overlooked. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely computationally expensive because it involves the simulation of thousands of tracks for each recorded event, which are performed each iteration of the Monte-Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the sheer number of events to analyse (in the hundreds of thousands) necessitated the use of parallel processing on a high performance computing cluster {\color{blue} is there a metric we can give here? -MPK}. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. 
%Most of the CPU time would be spent on simulating events that are not of interest.
The computationally expensive fitting procedure would be applied to every event, instead of the few percent of the events that are of interest for the analysis.
An unsupervised ML algorithm able to separate the data without {\em a priori} knowledge of the different types of events increases the efficiency of the analysis tremendously, and allows the downstream analysis to concentrate on the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data.}\label{tab:data sets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

@@

% In the Machine Learning community, it is not uncommon to publish packaged models with fitted parameters from image recognition contests. These models are trained on data sets with millions of images and classify between hundreds of distinct classes; one such is the imagenet data set. In their work, \citet{Kuchera2019} used the VGG16 architecture trained on imagenet to classify ${}^{46}$Ar data. In our work we also found the VGG16 architecture to capture the strongest event-separating representation for clustering purposes.

@@

Training high-performing neural networks from scratch often requires enormous data sets and computation time. However, it has been found that models which are trained at large scale will learn general features that are applicable to a variety of tasks. For example, large neural networks which are trained on the ImageNet data set \cite{Russakovsky2015} --- a diverse image classification task --- learn how to identify lines, edges, and other common shapes that are useful for numerous problems. Thus, it is common practice to initialize the convolutional layers of a network with the pre-trained weights learned from ImageNet (or some other large data set). The training process then only has to fine-tune the network for the specific task. Since we are building on prior knowledge in this case, learning becomes far more efficient, and better performance can often be achieved. \citet{Kuchera2019} used machine learning methods to classify the products of $^{46}$Ar reactions in the AT-TPC, and they found that a CNN initialized with ImageNet weights resulted in the most successful classification.

@@

%We begin by investigating the K-means clustering algorithm on the VGG16 latent space. 
% The user a VGG16 model that is pre-trained on the ImageNet data set and subsequently applied to the AT-TPC events without the classification layers. This creates a set of vectors $\boldsymbol{x} \in \R^{8192}$ as our new representation of the events. To cluster, we use the \lstinline{scikit-learn} implementation of the K-means++ algorithm with default parameters \cite{Pedregosa2011}.

@@

The performance is measured by comparing model predictions on the labelled subset of data (see table \ref{tab:data sets}).

@@

The results of the clustering runs are included in table \ref{tab:clstr_vgg}. We observe that we are able to attain near-perfect clustering on simulated data and that there is a sharp decline in performance as we add noise by moving to the filtered and raw data sets.

@@

{\color{blue} This section may contain info that you can move to the metrics section??}
In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we tabulate the assignments of samples relative to their ground truth labels. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in figures \ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. Purity is inferred by how much spread there is in the column between the ground truth labels. A high-quality cluster will, in addition to being pure, also capture most entries the class represented by the cluster. For example, consider the row corresponding to the proton class in figure \ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it. From this, we conclude that the proton cluster is a high quality, high purity cluster.

@@

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{Illustrating a sample of proton events from different K-means clusters from the filtered data set. Each row belongs to a single cluster corresponding to the filtered confusion matrix in figure \ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{Illustrating a sample of proton events from different K-means clusters from the raw data set. Each row belongs to a single cluster corresponding to the full confusion matrix in figure \ref{fig:clster_confmati_raw}}\label{fig:full_vgg_clster_repr}
\end{figure}

@@

\section{Conclusions and Perspectives}\label{sec{conclusion}}
% Begin edit by DB (6/17/2020)
The purpose of this study has been to explore the application of unsupervised learning algorithms to event identification from an active target detector. The necessity to identify events from raw data prior to full processing is becoming a major issue in the data analysis of detectors with complex responses such as the AT-TPC.
As shown by both avenues explored in this work, it is clear that there is significant potential to eventually achieve event classification using fully automated unsupervised methods.

@@

In summary, our study shows that unsupervised track classification with an implementation of the VGG16 + K-means approach is a viable solution. For future work, it is worth investigating whether the adaptation of  models like the MIXAE algorithm will allow better performance at no significant cost to consistency.  
%Especially considering the flexibility to use 3D input representations or auxiliary physical goals in the optimization procedure. 
The two examples of unsupervised machine learning methods studied in this work are a first encouraging step towards automated classification of events that could not only greatly reduce the resource cost of analysis, but also eventually boost the efficiency of the experiment by allowing post-trigger decisions based on such algorithm implemented in hardware. Much progress remains to be done before achieving these goals.
% End edit by DB (6/17/2020)
%From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. {\color{orange}[flip this to positive ]} It is also clear that there is significant potential in both avenues explored in this work. 
%n particular, the surprising ability of the K-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
%Another surprising facet of the K-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.

@@

\appendix  
\section{VGG16}\label{app:vgg}
The choice of the kernel size is based on the fact that a stacked $3 \times 3$ kernel is equivalent to larger kernels in terms of the receptive field of the output. Three $3 \times 3$ kernels with stride $1$ have a $7 \times 7$ receptive field, but the larger kernel has $81\%$ more parameters and only one non-linearity \cite{Simonyan2014}. Stacking the smaller kernels then contributes to a lower computational cost. Additionally, there is a regularizing effect from the lowered number of parameters and increased explanatory power from the additional non-linearities.
\begin{table}
\caption[vgg architectures]{Showing the details of the VGG network architectures. Network D trained on the ImageNet \cite{Russakovsky2015} data set the network known as  VGG16 and is what we use in this thesis.}\label{tab:vgg}
\includegraphics[width=\textwidth]{plots/vgg_architectures.png}
\end{table}

@@

To train the MIXAE clustering algorithm, we use the large simulated data set with $M=80000$ points, evenly distributed between proton- and carbon-events. The algorithm is trained on a subset of $60000$ of these samples, and we track performance on the remaining $20000$ events. Since there are then only three remaining hyperparameters we choose to perform a coarse grid-search as described in section \ref{sec:hyperparam_search_arch}.Finally, for these parameters we re-ran the algorithm $N=10$ times to investigate the stability of the algorithm.

@@

For the full data set the MIXAE hyperparameters converge to the same values as for the clean data:
